---
title: "R Notebook"
output: html_notebook
---

If you get lost, see especially [An Introduction to R](https://cran.r-project.org/doc/manuals/r-release/R-intro.pdf).

## Data types in R

### The assignment operator (<-)

- You might think it's like other programming languages, where `=` or `:=` is the assignment operator.
- You'd be wrong.
- R uses `<-` to indicate assignment.

```{r}

```

### Objects

### Vectors and lists

### Matrices

### Factors

### Data Frames

### The `names` function

## Reading input

### `file`

### `readLines`

### `read.csv`

### `url`

## Saving output

### `dput` and `dget`

### `dump` and `source`

## Subsetting

### `[]`

- Always returns an object of the same class as the original

### `[[]]`

- Extracts a single element from a list or data frame

### `$`

- Extracts elements by name --- kind of like `.` in some other languages

### Removing missing values

#### `complete.cases`

### `|>` (the pipe)

The pipe operator `|>` that passes its left-hand side as the first argument to the its right-hand side. In other words `[1 2 3] |> first()` is equivalent to `first([1 2 3])`.

You can try it for yourself below.

```{r}
foo <- 0:9

foo |> last() # should return the integer 9
```